### 2015-06-23 Manuscript/Missing_mammals-draft.tex
Abstract

\noindent{\Large \bf Morphological data is lacking for living mammals}
%Or: Missing cladistic data in living mammals - but I prefer the first one emphasizing on the availability of it. The data is not missing per se, it's just not coded!
% NC: sometimes the better title is one that answers the question rather than giving the topic. e.g., "Morphological data is lacking for living species" or similar
% TG: maybe worth considering the special issue as well? "Putting fossils in tree" so maybe something like "There is not enough morphological data for living mammals to branch fossils in the trees."
% TG: or some poetic... "Living and dead mammals: morphological data is lacking in living for combined analysis"

%Key words: Total Evidence method, data structure, phylogenetic, living, fossil, topology
\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^2$$^,$$^*$ and Natalie Cooper$^1$$^,$$^2$$^,$$^3$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
$^2$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland.\\
$^3$Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK.}\\
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{Zoology Building, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.}\\  
\vspace{1in}

@@

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------
% NC: Time to sort this out!
\newpage
\begin{abstract} % 200 words, all keywords written in abstract if possible.
Studying changes in global biodiversity through time and space is essential.
For that we need methods to combine both palaeontological and neontological data.
One promising method, the Total Evidence method, allows such thing but needs a lot of data.
Especially cladistic data from living taxa to allow the fossil taxa to accurately branch in the trees.
Despite two centuries of morphological studies on living taxa, scientists using and generating such data mainly focus on palaeontological data.
Therefore, even in well known groups such as mammal, there is a huge gap in our knowledge of cladistic data for living mammals.
In this study, using phylogenetic community structure methods, we quantify the availability of data in each mammalian order.
And maybe at the end of the paper we propose some discussion on how to improve all that (go in museums!).
\end{abstract}

\noindent ()\\

@@

% NC: If there are word count problems we can ditch the last two sentences.
%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------
\section{Materials and Methods}
% NC: Need to add the dates, or date ranges, of your data collection below instead of @
\subsection{Data collection and standardisation}
We downloaded all cladistic matrices containing any living and/or fossil mammal taxa from three major public databases (accessed @th of July 2015): Morphobank \citep[\texttt{http://www.morphobank.org/};][]{morphobank}, Graeme Lloyd's website (\texttt{graemetlloyd.com/matrmamm.html}) and Ross Mounce's GitHub repository (\texttt{https://github.com/rossmounce/cladistic-data}).
We also performed a systematic Google Scholar search (accessed @th of July 2015) for matrices that were not uploaded to these databases (see Supplementary Materials for a detailed description of the search procedure).
In total, we downloaded 286 matrices containing a total of 11010 operational taxonomic units (OTUs) of which 5228 were unique. % Number of unique OTUs (including fossils)
In this study, we refer to OTUs rather than species since the entries in the downloaded matrices were not standardised and ranged from specific individual specimen names (i.e. the name of a collection item) to the family-level.
Where possible, we considered OTUs at their lowest valid taxonomic level (i.e. species) but some OTUs were only valid at a higher taxonomic level (e.g. genus or family).
Therefore for some orders, we sampled more genera than species (Table \ref{Table_morpho_taxa_proportion}).

@@

For orders with \textless 100\% cladistic data coverage at any taxonomic level, we investigated whether the available cladistic data was (i) randomly distributed, (ii) overdispersed or (iii) clustered, with respect to phylogeny, using two metrics from community phylogenetics: the Nearest Taxon Index \citep[NTI;][]{webb2002phylogenies} and the Net Relatedness Index \citep[NRI;][]{webb2002phylogenies}. 
NTI is most sensitive to clustering or overdispersion near the tips, whereas NRI is more sensitive to clustering or overdispersion across the whole phylogeny \citep{Cooper2008}. 
% NC: I moved this back up to keep it shorter.
Both metrics were calculated using the \texttt{picante} package in R \citep{picante,R}.

NTI \citep{webb2002phylogenies} is based on mean nearest neighbour distance (MNND) and is calculated as follows:
  \begin{equation}
    NTI=-\left(\frac{\overline{MNND}_{obs}-\overline{MNND}_{n}}{\sigma(MNND_{n})}\right)
  \end{equation}
where $\overline{MNND}_{obs}$ is the observed mean distance between each of $n$ taxa with cladistic data and its nearest neighbour with cladistic data in the phylogeny, 
% NC: Argh still wrong definition of MNND, I've fixed it myself. It's PAIRWISE!!!
$\overline{MNND}_{n}$ is the mean of 1000 mean MNND between $n$ randomly drawn taxa, and $\sigma(MNND_{n})$ is the standard deviation of these 1000 random MNND values.
NRI is similar but is based on mean phylogenetic distance (MPD) as follows:
  \begin{equation}
    NRI=-\left(\frac{\overline{MPD}_{obs}-\overline{MPD}_{n}}{\sigma(MPD_{n})}\right)
  \end{equation}
where $\overline{MPD}_{obs}$ is the observed mean phylogenetic distance of the tree containing only the $n$ taxa with cladistic data, $\overline{MPD}_{n}$ is the expected random MPD for $n$ taxa estimated by calculating the MPD from $n$ taxa randomly drawn from the phylogeny and repeated 1000 times, and $\sigma(MPD_{n})$ is the standard deviation of the 1000 random MPD values. % NC: This can be shortened a bit but it's only gonna save you 50 words max.
Negative NTI and NRI values show that the focal taxa are more overdispersed across the phylogeny than expected by chance, and positive values reflect significant clustering.

@@

\section{Results}
% NC: Need to replace these @?
Across the @126 cladistic matrices we extracted, @22 out of 28 mammalian orders have low coverage (\textless 25\% of species with cladistic data) and @six have high coverage (\textgreater 75\% of species with cladistic data) at the species-level.
At the genus-level, @three orders have low coverage and @12 have high coverage, and at the family-level, no orders have low coverage and @23 have high coverage (Table \ref{Table_morpho_taxa_proportion}).

@@

The number of living mammalian taxa with no available cladistic data was surprisingly high at the species-level: only @six out of @28 orders have a high coverage of taxa with available cladistic data (and two of the 28 orders are monospecific!).
This high coverage threshold of 75\% of taxa with available cladistic data represents the minimum amount of data required before missing data has a significant effect on the topology of Total Evidence trees \citep{GuillermeCooper}.
Beyond this threshold, there is considerable displacement of wildcard taxa (\textit{sensu} \citep{kearneyfragmentary2002}) and decreases in clade conservation \citep{GuillermeCooper}.
Therefore we expect a high probability of topological artefacts for the placement of fossil taxa at the species-level in most mammalian orders.
However, data coverage seems to be less of an issue at higher taxonomic levels (i.e. genus- and family-level).
This point is important from a practical point of view because of the slight discrepancy between the neontological and palaeontological concept of species.
While neontological species are described using morphology, genetic distance, spatial distribution and even behaviour, palaeontological species can be based only on morphological, spatial and temporal data \citep[e.g.][]{ni2013oldest}.
Because of this, most palaeontological studies are using the genus as their smallest OTU \citep[e.g.][]{ni2013oldest,O'Leary08022013}.
Thus data availability at the genus-level in living mammals should be our primary concern when aiming to build phylogenies of living and fossil taxa.

When only a few species with cladistic data are available, the ideal scenario is for them to be phylogenetically overdispersed (i.e. that there is data for at least every sub-clade) to maximize the possibilities of a fossil branching from the right clade.
The second best scenario is that species with cladistic data are randomly distributed across the phylogeny. 
In this scenario we expect no special bias in the placement of the fossil \citep{GuillermeCooper}, it is therefore encouraging that for most orders, species with cladistic data were randomly distributed across the phylogeny of each order.
The worst case scenario for fossil placement is that species with cladistic data are phylogenetically clustered. 
In this situation we expect two major biases to occur: first, the fossil will not be able to branch within a clade containing no data, and second, the fossil will have a higher probability, at random, of branching within the clade containing most of the available data.
This means that fossils with uncertain phylogenetic affinities (\textit{incertae sedis}) will have a higher probability of branching within the most sampled clade just by chance.
Our results suggest that this may be an issue, at the genus-level, in Carnivora, Cetartiodactyla, Chiroptera and Soricomorpha. 
For example, a Carnivora fossil will be unable to branch in the Herpestidae that has no species with cladistic data, and will also have more chance to branch, randomly, within the Canidae clade than any other clade in Carnivora (Figure \ref{Figure_example_coverage}B).
Thus, in Total Evidence trees, placements of some carnivoran fossils should considered with caution. 
% NC: Trying not to piss off GraHam - really it's weird stuff appearing in Carnivora that we are concerned by, not all Carnivora
In this study, we treated all cladistic matrices as equal in a similar way to molecular matrices. 
For example, if matrix A contained 100 characters for taxa X and Y, and matrix B contained 50 different characters for taxa X and Z, we assumed that both matrices can be combined in a supermatrix containing 150 independent characters for taxon X, 100 for taxon Y and 50 for taxon Z.
Unfortunately, cladistic data cannot always be treated in this way because some characters may overlap.
For example, if matrix A has a character coding for the shape of a particular morphological feature and matrix B has a character coding for the presence of this same morphological feature and a second character coding for its shape, then these three characters are non-independent compound characters \citep{Brazeau2011}.
However, in reasonably sized matrices (\textgreater 100 characters; \citealp{GuillermeCooper,harrisonamong-character2014}) it is more likely that a number of characters are consistently conserved among the different matrices and thus easily combinable.
For example, within the Primate cladistic literature, the character \textit{p7} - the size of the $4^{th}$ lower premolar paraconid - has been used consistently for \textgreater 15 years \citep[e.g.][]{ross1998phylogenetic,marivaux2005anthropoid,kay2008anatomy,boyer2010astragalar,ni2013oldest} and can therefore be combined among the matrices.
A conservative approach to avoid compound characters would be to select only the most recent matrix for each group, but this would result in the loss of a lot of data.

Despite the absence of good cladistic data coverage for living mammals, the Total Evidence methods still seems to be the most promising way of combining living and fossil data for macroevolutionary analyses. 
Following the recommendations in \citep{GuillermeCooper}, we need to code cladistic characters for as many living species possible. 
Fortunately, data for living mammals is usually readily available in natural history collections, therefore, we propose that an increased effort be put into coding morphological characters from living species, possibly by engaging in collaborative data collection projects through web portals such as \textit{Morphobank} \citep{morphobank}.
% NC: Still feels like it needs one final sentence to round it all off. Many something like below, but less crappy:
Such an effort would be valuable not only to phylogeneticists, but also to any researcher interested in the diversity of life on Earth.

>>>>

\noindent{\Large \bf Morphological data is lacking for living mammals}
%Or: Missing cladistic data in living mammals - but I prefer the first one emphasizing on the availability of it. The data is not missing per se, it's just not coded!
% NC: sometimes the better title is one that answers the question rather than giving the topic. e.g., "Morphological data is lacking for living species" or similar
% TG: maybe worth considering the special issue as well? "Putting fossils in tree" so maybe something like "There is not enough morphological data for living mammals to branch fossils in the trees."
% TG: or some poetic... "Living and dead mammals: morphological data is lacking in living for combined analysis"
\bigskip

\noindent {\normalsize \sc Thomas Guillerme$^1$$^,$$^2$$^,$$^*$ and Natalie Cooper$^1$$^,$$^2$$^,$$^3$}\\
\noindent {\small \it 
$^1$School of Natural Sciences, Trinity College Dublin, Dublin 2, Ireland.\\
$^2$Trinity Centre for Biodiversity Research, Trinity College Dublin, Dublin 2, Ireland.\\ %Should we remove that one for word count?
$^3$Department of Life Sciences, Natural History Museum, Cromwell Road, London, SW7 5BD, UK.}\\
\end{center}
\medskip
\noindent{*\bf Corresponding author.} \textit{Zoology Building, Trinity College Dublin, Dublin 2, Ireland; E-mail: guillert@tcd.ie; Fax: +353 1 6778094; Tel: +353 1 896 2571.}\\  
\vspace{1in}

@@

%---------------------------------------------
%
%       ABSTRACT
%
%---------------------------------------------
% NC: Time to sort this out!
\newpage
\begin{abstract} % 200 words, all keywords written in abstract if possible. TG: That's 195 words for now.
Combining living and fossil in the same analysis data is crucial for studying changes in global biodiversity through time.
One method allowing to combine this data is the Total Evidence method that uses both molecular data for living species and morphological data for both living and fossil species.
With this method, a good overlap of morphological data between living and fossil taxa is crucial for accurately inferring the phylogenies’ topology.

Since the advent of DNA, molecular data has become easily and widely available.
However, despite two centuries of morphological studies, scientists using and generating such data mainly focus on palaeontological data.
Therefore, there is a gap in our knowledge of neontological morphological data even in well studied groups such as mammals.

In this study, we quantify the morphological data available for living mammal taxa.
We then analyse the structure of the available data by testing if it is clustered or evenly spread across the phylogeny.
We found that 78\% of mammalian orders have less than 25\% data available at the species level.
However, we found that the available is often randomly distributed among these orders apart from six of them where the data is clustered
\end{abstract}

%\noindent Key words: Total Evidence method, data structure, phylogenetic, fossil, topology\\ Excluded from word count.

@@

% NC: If there are word count problems we can ditch the last two sentences.
%---------------------------------------------
%
%       METHODS
%
%---------------------------------------------
\section{Materials and Methods}
\subsection{Data collection and standardisation}
We downloaded all cladistic matrices containing any living and/or fossil mammal taxa from three major public databases (accessed 10th of June 2015): Morphobank \citep[\texttt{http://www.morphobank.org/};][]{morphobank}, Graeme Lloyd's website (\texttt{graemetlloyd.com/matrmamm.html}) and Ross Mounce's GitHub repository (\texttt{https://github.com/rossmounce/cladistic-data}).
We also performed a systematic Google Scholar search (accessed 11th of June 2015) for matrices that were not uploaded to these databases (see Supplementary Materials for a detailed description of the search procedure).
In total, we downloaded 286 matrices containing a total of 11010 operational taxonomic units (OTUs) of which 5228 were unique. % Number of unique OTUs (including fossils)
In this study, we refer to OTUs rather than species since the entries in the downloaded matrices were not standardised and ranged from specific individual specimen names (i.e. the name of a collection item) to the family-level.
Where possible, we considered OTUs at their lowest valid taxonomic level (i.e. species) but some OTUs were only valid at a higher taxonomic level (e.g. genus or family).
Therefore for some orders, we sampled more genera than species (Table \ref{Table_morpho_taxa_proportion}).

@@

For orders with \textless 100\% cladistic data coverage at any taxonomic level, we investigated whether the available cladistic data was (i) randomly distributed, (ii) overdispersed or (iii) clustered, with respect to phylogeny, using two metrics from community phylogenetics: the Nearest Taxon Index \citep[NTI;][]{webb2002phylogenies} and the Net Relatedness Index \citep[NRI;][]{webb2002phylogenies}. 
NTI is most sensitive to clustering or overdispersion near the tips, whereas NRI is more sensitive to clustering or overdispersion across the whole phylogeny \citep{Cooper2008}. 
Both metrics were calculated using the \texttt{picante} package in R \citep{picante,R}.

NTI \citep{webb2002phylogenies} is based on mean nearest neighbour distance (MNND) and is calculated as follows:
  \begin{equation}
    NTI=-\left(\frac{\overline{MNND}_{obs}-\overline{MNND}_{n}}{\sigma(MNND_{n})}\right)
  \end{equation}
where $\overline{MNND}_{obs}$ is the observed mean distance between each of $n$ taxa with cladistic data and its nearest neighbour with cladistic data in the phylogeny, 
$\overline{MNND}_{n}$ is the mean of 1000 mean MNND between $n$ randomly drawn taxa, and $\sigma(MNND_{n})$ is the standard deviation of these 1000 random MNND values.
NRI is similar but is based on mean phylogenetic distance (MPD) as follows:
  \begin{equation}
    NRI=-\left(\frac{\overline{MPD}_{obs}-\overline{MPD}_{n}}{\sigma(MPD_{n})}\right)
  \end{equation}
where $\overline{MPD}_{obs}$ is the observed mean phylogenetic distance of the tree containing only the $n$ taxa with cladistic data, $\overline{MPD}_{n}$ is the expected random MPD for $n$ taxa estimated by calculating the MPD from $n$ taxa randomly drawn from the phylogeny and repeated 1000 times, and $\sigma(MPD_{n})$ is the standard deviation of the 1000 random MPD values. % NC: This can be shortened a bit but it's only gonna save you 50 words max.
Negative NTI and NRI values show that the focal taxa are more overdispersed across the phylogeny than expected by chance, and positive values reflect significant clustering.

@@

\section{Results}
Across the 126 cladistic matrices we extracted, 22 out of 28 mammalian orders have low coverage (\textless 25\% of species with cladistic data) and six have high coverage (\textgreater 75\% of species with cladistic data) at the species-level.
At the genus-level, three orders have low coverage and 12 have high coverage, and at the family-level, no orders have low coverage and 23 have high coverage (Table \ref{Table_morpho_taxa_proportion}).

@@

The number of living mammalian taxa with no available cladistic data was surprisingly high at the species-level: only six out of 28 orders have a high coverage of taxa with available cladistic data (and two of the 28 orders are monospecific!).
This high coverage threshold of 75\% of taxa with available cladistic data represents the minimum amount of data required before missing data has a significant effect on the topology of Total Evidence trees \citep{GuillermeCooper}.
Beyond this threshold, there is considerable displacement of wildcard taxa (\textit{sensu} \citep{kearneyfragmentary2002}) and decreases in clade conservation \citep{GuillermeCooper}.
Therefore we expect a high probability of topological artefacts for the placement of fossil taxa at the species-level in most mammalian orders.
However, data coverage seems to be less of an issue at higher taxonomic levels (i.e. genus- and family-level).
This point is important from a practical point of view because of the slight discrepancy between the neontological and palaeontological concept of species.
While neontological species are described using morphology, genetic distance, spatial distribution and even behaviour, palaeontological species can be based only on morphological, spatial and temporal data \citep[e.g.][]{ni2013oldest}.
Because of this, most palaeontological studies are using the genus as their smallest OTU \citep[e.g.][]{ni2013oldest,O'Leary08022013}.
Thus data availability at the genus-level in living mammals should be our primary concern when aiming to build phylogenies of living and fossil taxa.

When only a few species with cladistic data are available, the ideal scenario is for them to be phylogenetically overdispersed (i.e. that there is data for at least every sub-clade) to maximize the possibilities of a fossil branching from the right clade.
The second best scenario is that species with cladistic data are randomly distributed across the phylogeny. 
In this scenario we expect no special bias in the placement of the fossil \citep{GuillermeCooper}, it is therefore encouraging that for most orders, species with cladistic data were randomly distributed across the phylogeny of each order.
The worst case scenario for fossil placement is that species with cladistic data are phylogenetically clustered. 
In this situation we expect two major biases to occur: first, the fossil will not be able to branch within a clade containing no data, and second, the fossil will have a higher probability, at random, of branching within the clade containing most of the available data.
This means that fossils with uncertain phylogenetic affinities (\textit{incertae sedis}) will have a higher probability of branching within the most sampled clade just by chance.
Our results suggest that this may be an issue, at the genus-level, in Carnivora, Cetartiodactyla, Chiroptera and Soricomorpha. 
For example, a Carnivora fossil will be unable to branch in the Herpestidae that has no species with cladistic data, and will also have more chance to branch, randomly, within the Canidae clade than any other clade in Carnivora (Figure \ref{Figure_example_coverage}B).
Thus, in Total Evidence trees, placements of some carnivoran fossils should considered with caution. 
% NC: Trying not to piss off GraHam - really it's weird stuff appearing in Carnivora that we are concerned by, not all Carnivora. TG: I think it's mainly Graham's work that focuses on canids.
In this study, we treated all cladistic matrices as equal in a similar way to molecular matrices. 
For example, if matrix A contained 100 characters for taxa X and Y, and matrix B contained 50 different characters for taxa X and Z, we assumed that both matrices can be combined in a supermatrix containing 150 independent characters for taxon X, 100 for taxon Y and 50 for taxon Z.
Unfortunately, cladistic data cannot always be treated in this way because some characters may overlap.
For example, if matrix A has a character coding for the shape of a particular morphological feature and matrix B has a character coding for the presence of this same morphological feature and a second character coding for its shape, then these three characters are non-independent compound characters \citep{Brazeau2011}.
However, in reasonably sized matrices (\textgreater 100 characters; \citealp{GuillermeCooper,harrisonamong-character2014}) it is more likely that a number of characters are consistently conserved among the different matrices and thus easily combinable.
For example, within the Primate cladistic literature, the character \textit{p7} - the size of the $4^{th}$ lower premolar paraconid - has been used consistently for \textgreater 15 years \citep[e.g.][]{ross1998phylogenetic,marivaux2005anthropoid,kay2008anatomy,boyer2010astragalar,ni2013oldest} and can therefore be combined among the matrices.
A conservative approach to avoid compound characters would be to select only the most recent matrix for each group, but this would result in the loss of a lot of data.

Despite the absence of good cladistic data coverage for living mammals, the Total Evidence methods still seems to be the most promising way of combining living and fossil data for macroevolutionary analyses. 
Following the recommendations in \citep{GuillermeCooper}, we need to code cladistic characters for as many living species possible. 
Fortunately, data for living mammals is usually readily available in natural history collections, therefore, we propose that an increased effort be put into coding morphological characters from living species, possibly by engaging in collaborative data collection projects through web portals such as \textit{Morphobank} \citep{morphobank}.
% NC: Still feels like it needs one final sentence to round it all off. Many something like below, but less crappy:
Such an effort would be valuable not only to phylogeneticists, but also to any researcher focusing understanding macroevolutionary patterns and processes.%interested in the diversity of life on Earth.
